Investigating the alternative prey hypothesis with the POMP framework

### 2024-04-24 unblinded.Rmd
Add files via upload

---
title: "Investigating the alternative prey hypothesis with the POMP framework"
author: "Pete, Sizhuang, and Heyuan"
date: "`r Sys.Date()`"
csl: https://raw.githubusercontent.com/citation-style-language/styles/master/cell-numeric.csl
bibliography: references.bib 
output: 
  bookdown::html_document2:
    code_folding: hide
    theme: cosmo
    toc: yes
    toc_float:
      collapsed: true
---

![Female Willow Ptarmigan captured by [Bryce W. Robinson](https://ornithologi.com/2015/06/30/in-context-mid-summer-willow-ptarmigan-behavior-and-appearance/).](./assets/ptarmigan.jpg)

@@

```{r libraries, echo=FALSE, message=FALSE, warning=FALSE}
library(readxl)
library(ggplot2)
library(knitr)
library(forecast)
library(dplyr)
library(cowplot)
library(pomp)
plan(multisession)
library(doParallel)
library(doRNG)
library(foreach)
library(doFuture)
```

```{r data_table, echo=FALSE}
df <- read_excel("./data/bird_data.xlsx")
df |>
  rename(R = "peak rodent year") |>
  rename(NAO = "NAO May_Jun_Jul") |>
  rename(logCPUE = "log.CPUE") |>
  mutate(R = ifelse(R == "no", 0, 1)) -> data

@@

ARMA model can't be performed on the non-stationary time series data. The KPSS test is implemented for a stationarity check.
```{r stationary, echo=FALSE}
# Kwiatkowski-Phillips-Schmidt-Shin (KPSS) test
kpss_result <- tseries::kpss.test(ts_cpue, null = "Level")
print(kpss_result)

# Differencing the time series once
start_year <- min(data$year)
end_year <- max(data$year)
frequency <- 1  
ts_cpue <- ts(data$logCPUE, start=start_year, end=end_year, frequency=frequency)
ts_diff <- diff(ts_cpue, differences = 1)
print(tseries::kpss.test(ts_diff,null = "Level"))
```

@@

![Conditional independence graph of POMP from A. King's Notes](./assets/state_space.png)

@@

```{r, echo=FALSE, eval=FALSE}
mif2(
  mod, 
  Np = 1000, 
  Nmif = 50, 
  rw.sd = rw_sd(
    alpha = 0.02, Beta = 0.02, gamma = 0.02, 
    a = 0.02, b = 0.02, c = 0.02, 
    sigma_obs = 0.02, logF_0 = ivp(0.1), logB_0 = ivp(0.1), 
    sigmaF = 0.02, sigmaB = 0.02, logRho = 0.02
  ),
  cooling.fraction.50 = 0.5,
  params = param_guess
) -> mif2_out

@@

The estimated log-likelihood is -205 with standard error of 3.14.

@@

@@

It estimated log likelihood of -134 which is larger than a single execution of `mif2` function, but with a larger standard error of 8.2.
![Model parameters over iterations](./assets/local_search_1.png)
From the above plot, the likelihood over iterations quickly rises and then levels off. Towards the later iterations, the changes in parameter estimates become smaller, indicating that the search for the maximum likelihood is flattening out. This can be a sign that the search is approaching a local maximum, but the fact that the changes are small but not zero implies that there might still be some uncertainty. But not all parameters show signs of convergence, such as $log(B_0), \sigma_F$.

@@

```{r}
pairs(~loglik+a+alpha+b+Beta+c+gamma+logB_0+logF_0+logRho+sigmaB+sigmaF,data=results_local,pch=16)
```

![](./assets/local_search_2.png)
A scattered or cloud-like pattern indicates little to no apparent linear correlation.

@@

```{r}
set.seed(2062379496)

@@

```{r, echo=FALSE}
results <- read.csv('/Users/ruojunliu/Desktop/bird_params_middle.csv')
print(paste('The maximum likelihood from global search for our model is', max(results$loglik), 
            'with a standard error of ', results$loglik.se[which.max(results$loglik)]))

@@

As a result, he log-likelihood produced an ARIMA(0,1,5) model is better than that from a local search of a POMP model, it implies that the ARMA model is fitting the data better according to this metric

>>>>

---
title: "Investigating the alternative prey hypothesis with the POMP framework"
author: "Pete, Sizhuang, and Heyuan"
date: "`r Sys.Date()`"
csl: https://raw.githubusercontent.com/citation-style-language/styles/master/cell-numeric.csl
bibliography: /Users/ruojunliu/Desktop/references.bib 
output: 
  bookdown::html_document2:
    code_folding: hide
    theme: cosmo
    toc: yes
    toc_float:
      collapsed: true
---

![Female Willow Ptarmigan captured by [Bryce W. Robinson](https://ornithologi.com/2015/06/30/in-context-mid-summer-willow-ptarmigan-behavior-and-appearance/).](/Users/ruojunliu/Desktop/STATS 531 - Time Series/pomp_final/assets/ptarmigan.jpg)

@@

```{r libraries, echo=FALSE, message=FALSE, warning=FALSE}
library(readxl)
library(ggplot2)
library(knitr)
library(forecast)
library(dplyr)
library(cowplot)
library(pomp)
library(doParallel)
library(doRNG)
library(foreach)
library(doFuture)
```

```{r data_table, echo=FALSE}
df <- read_excel("/Users/ruojunliu/Desktop/STATS 531 - Time Series/pomp_final/data/bird_data.xlsx")
df |>
  rename(R = "peak rodent year") |>
  rename(NAO = "NAO May_Jun_Jul") |>
  rename(logCPUE = "log.CPUE") |>
  mutate(R = ifelse(R == "no", 0, 1)) -> data

@@

ARMA model can't be performed on the non-stationary time series data. The KPSS test is implemented for a stationarity check.
```{r stationary, echo=FALSE}
start_year <- min(data$year)
end_year <- max(data$year)
frequency <- 1  
ts_cpue <- ts(data$logCPUE, start=start_year, end=end_year, frequency=frequency)

# Kwiatkowski-Phillips-Schmidt-Shin (KPSS) test
kpss_result <- tseries::kpss.test(ts_cpue, null = "Level")
print(kpss_result)

# Differencing the time series once
ts_diff <- diff(ts_cpue, differences = 1)
print(tseries::kpss.test(ts_diff,null = "Level"))
```

@@

![Conditional independence graph of POMP from A. King's Notes](/Users/ruojunliu/Desktop/STATS 531 - Time Series/pomp_final/assets/state_space.png)

@@

```{r, echo=FALSE}
mif2(
  mod, 
  Np = 1000, 
  Nmif = 50, 
  rw.sd = rw_sd(
    alpha = 0.02, Beta = 0.02, gamma = 0.02, 
    a = 0.02, b = 0.02, c = 0.02, 
    sigma_obs = 0.02, logF_0 = ivp(0.1), logB_0 = ivp(0.1), 
    sigmaF = 0.02, sigmaB = 0.02, logRho = 0.02
  ),
  cooling.fraction.50 = 0.5,
  params = param_guess
) -> mif2_out

@@

The estimated log-likelihood is **-205** with standard error of 3.14.

@@

pairs(~loglik+a+alpha+b+Beta+c+gamma+logB_0+logF_0+logRho+sigmaB+sigmaF,data=results_local,pch=16)

@@

It estimated log likelihood of **-134** which is larger than a single execution of `mif2` function, but with a larger standard error of 8.2.
![Model parameters over iterations](/Users/ruojunliu/Desktop/STATS 531 - Time Series/pomp_final/assets/local_search_1.png)

From the above plot, the likelihood over iterations quickly rises and then levels off. Towards the later iterations, the changes in parameter estimates become smaller, indicating that the search for the maximum likelihood is flattening out. This can be a sign that the search is approaching a local maximum, but the fact that the changes are small but not zero implies that there might still be some uncertainty. But not all parameters show signs of convergence, such as $log(B_0), \sigma_F$.

@@

```{r, echo=FALSE, eval=FALSE}
#pairs(~loglik+a+alpha+b+Beta+c+gamma+logB_0+logF_0+logRho+sigmaB+sigmaF,data=results_local,pch=16)
```

![](/Users/ruojunliu/Desktop/STATS 531 - Time Series/pomp_final/assets/local_search_2.png)

A scattered or cloud-like pattern indicates little to no apparent linear correlation.

@@

```{r, echo=FALSE, eval=FALSE}
set.seed(2062379496)

@@

```{r, echo=FALSE, eval=FALSE}
results <- read.csv('/Users/ruojunliu/Desktop/bird_params_middle.csv')
print(paste('The maximum likelihood from global search for our model is', max(results$loglik), 
            'with a standard error of ', results$loglik.se[which.max(results$loglik)]))

@@

As a result, he log-likelihood produced an ARIMA(0,1,5) model is better than that from a local search of a POMP model, it implies that the ARMA model is fitting the data better according to this metric.
